Fusing Filters with Integer Linear Programming

### 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex
size inference: drop superceeded text
pl. of scheme is schemes, wheraas pl. of schema is schemata. scheme and schema are different things (I think)
Redo iteration size section, use \bot to indicate an unknown size

% Our formulation of size inference is very similar to type inference, particularly the exposition of HM(X). First, constraints are generated for the bindings, such as equality among two sizes, the conjunction of two constraints, and existentials. If these constraints can be solved, the equality constraints are used to group the sizes into equivalence classes. Otherwise if the constraints are unable to be solved, the program cannot be assured to require no runtime size checks and will not be fused by our system. After the constraints are generated and solved, each combinator is given an iteration size. Two loops with the same iteration size can be fused together.

% While it \emph{is} technically possible to fuse loops of different sizes, it requires extra complexity to find the maximum of the loop sizes, and additional branches are required to only execute the smaller loop fewer times. For simplicity, we do away with this added complexity and only support fusion of equal sized loops. \emph{Size inference} is performed on the combinators to infer as much information as possible about the sizes of the resulting loops.

% A more in-depth analysis has also been explored by Jay~\cite{jay1996shape} in the form of shape inference, which is built from primitives such as @cons@ and @nil@. Since we have fewer primitives and simpler goals, our size inference does not need to be as complicated as Jay's shape inference.

% The $@map@_n$ combinators require all input arrays to be the same size, and their output is the same. Since @fold@s do not produce arrays, they have no constraints. @gather@ takes two arrays; the data and the indices. Its output size is the size of the indices. The size of a @filter@'s output is most interesting; the exact size is not known until after it has been executed, only that it is less than or equal to the input size. Similarly, the size of @external@ outputs is not known at all, and thus cannot be constrained.

@@

\begin{tabbing}
MMMMMMMM \= MM  \= MM \= MMMMMM \= \kill
\textbf{Size Constraint}
\> $C$      \> @::=@  \> $true$               \> (trivially true)      \\
\>          \> $~|$   \> $\tau = \tau$        \> (equality constraint) \\
\>          \> $~|$   \> $C \wedge C$         \> (conjunction)
\end{tabbing}

@@

A size may either be a variable $k$, or a cross product of two sizes. We use the latter to represent the result size of the @cross@ operator discussed in the previous section. Constraints maybe either be trivially $true$, an equality $\tau = \tau$, or a conjunction of two constraints $C \wedge C$. We refer to the trivially true and equality constraints as \emph{atomic constraints}. Size schemes relate the sizes of each input and output array. For example, the size scheme for the @normalize@ example from Figure~\ref{f:normalize-clusterings} is as follows:
$$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

We write $:_s$ to distinguish size schemata from type schemata.

@@

% As an example of constraint generation, here is the @normalize2@ program from earlier.
% \begin{tabbing}
% @MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
% @normalize2 us@                     \> $\exists k_{us}.$      \\
% @ = let sum1 = fold   (+) 0 us@     \>                      \\
% @       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
% @       sum2 = fold   (+) 0 gts@    \> \\
% @       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
% @       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
% @   in (nor1, nor2)@                \> $true$ \\
% \end{tabbing}
% This constraint is valid, with the equivalence classes being:
% \eqclasses{
%     \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
%     \eqclass{k_{gts}}{k_{gts}}
% }

@@

% -----------------------------------------------------------------------------
\subsection{Iteration size}
After the constraints are validated and equivalence classes generated, each combinator is assigned a $size$ as an iteration size -- the number of iterations in the loop required to generate the output.
It is important to note that for filters, the iteration size is not the output size, but is instead the size of the input.
The output size of a filter is, however, a \emph{subsize} of the input's size, as not only is it known to be less than or equal to its input in size, it is also generated depending on the input.
The iteration sizes, $\tau_n$, are used to check whether two loops may be fused.
Any two iteration sizes in the same equivalence class are the same size, and so are fusible.
The difference from previous work is that loops of different iteration sizes \emph{can} be fused, if one is a subsize of the other, and the subsize's \emph{generator} is fused together it as well.
Basically, operations on filtered data can be fused with operations on the original data, if it is fused with the filter as well.
External computations are treated separately, as they cannot be fused with any other nodes.

\begin{tabbing}
MMMMM       \= MM \= MMMMMMMMMMM \= \kill
$T$          \> @::=@ \> $\tau$                                  \> (known size) \\
             \> $~|$  \> @external@                              \> (external and unfusible) \\
\end{tabbing}

Once the constraints are solved, known to be valid, and sorted into equivalence classes, each combinator is assigned a size.
Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
This is because, in order to produce the filtered output, all elements of the input $n$ must be considered.

\begin{tabbing}
MM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
$\tau$  \>$::$\> $binds \rightarrow name \rightarrow T$ \\
\\
$\tau_{bs,o}$    
            \> $|$ \> $o = @fold@~f~n$      \> $\in bs$ \> $=$ \> $k_n$ \\
            \> $|$ \> $o = @map@_n~f~ns$    \> $\in bs$ \> $=$ \> $k_o$ \\
            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $k_n$ \\
            \> $|$ \> $o = @generate@~s~f$  \> $\in bs$ \> $=$ \> $k_o$ \\
            \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $k_i$ \\
            \> $|$ \> $o = @cross@~a~b$     \> $\in bs$ \> $=$ \> $k_a \times k_b$ \\
            \> $|$ \> $o = @external@~ins$  \> $\in bs$ \> $=$ \> $@external@$ \\
\end{tabbing}

After the generated equality constraints are solved, and sizes are grouped into equivalence classes, combinators with iteration sizes in the same equivalence class may be fused together if there are no fusion-preventing dependencies between them.

>>>>

@@

\begin{tabbing}
MMMMMMMM \= MM  \= MM \= MMMMMM \= \kill
\textbf{Size Constraint}
\> $C$      \> @::=@  \> $true$               \> (trivially true)      \\
\>          \> $~|$   \> $k = \tau$           \> (equality constraint) \\
\>          \> $~|$   \> $C \wedge C$         \> (conjunction)
\end{tabbing}

@@

A size may either be a variable $k$, or a cross product of two sizes. We use the latter to represent the result size of the @cross@ operator discussed in the previous section. Constraints maybe either be trivially $true$, an equality $\tau = \tau$, or a conjunction of two constraints $C \wedge C$. We refer to the trivially true and equality constraints as \emph{atomic constraints}. Size schemes relate the sizes of each input and output array. For example, the size scheme for the @normalize@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
$$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

We write $:_s$ to distinguish size schemes from type schemes.

@@

@@

% -----------------------------------------------------------------------------
\subsection{Iteration size}
After the inferring the size of each array variable, each operator is assigned an \emph{iteration size}, which is the number of iterations needed in the loop which evaluates that operator. For @filter@ and other size changing operators, the iteration and result sizes are necessarily different. For such an operator, we say that the result size is a \emph{descendant} of the iteration size. Conversely, the iteration size is a \emph{parent} of the result size.

This descendant--parent size relation is transitive, so if we filter an array, then filter it again, the size of the result is a descendant of the iteration size of the initial filter. This relation arises naturally from Data Flow Fusion~\cite{lippmeier2013flow}, as such an operation would be compiled into a single loop --- with an iteration size identical to the size of the input array, and containing two nested if-expressions to perform the two layers of filtering.

Iteration sizes are used to decide which operators can be fused with each other. As in prior work, operators with the same iteration size can be fused. However, in our system we also allow operators of different iteration sizes to be fused, provided those sizes are descendants of the same parent size.

We use $T$ to range over iteration sizes, and write $\bot$ for the case where the iteration size is unknown. The $\bot$ size is needed to handle the @external@ operator, as in general we cannot know its true iteration size, and it cannot be fused with any other operator.

\begin{tabbing}
MMMMMMMM \= MM       \= MM \= MMMM \= \kill
\textbf{Iteration Size}
 \> $T$         \> @::=@  \> $\tau$        \> (known size) \\
 \>             \> $~|$   \> $\bot$     \> (unknown size) \\
\end{tabbing}

Once the size constraints have been solved, we can use the following function to compute the iteration size of each binding. In the definition we use the syntax $\Gamma(xs)$ to find the $xs : k$ element in the environment $\Gamma$ and return the associated size $k$. Similarly, we use the syntax $C(k)$ to find the corresponding $k = \tau$ constraint in $C$ and return the associated size type $\tau$.

\begin{tabbing}
MMMMM \= M \= MMMMMMMMMM \= MM \= \kill
$\iiter_{\Gamma,C}$  
        \>$:$\> $bind \rightarrow T$ 
\\[1ex]
$\iiter_{\Gamma,C}$
        \> $|$  \> $(z~ = @fold@~ f~xs)$     
                \> $=$ \> $C(\Gamma(xs))$ 
\\
        \> $|$  \> $(ys = @map@_n~f~\overline{xs})$
                \> $=$ \> $C(\Gamma(ys))$ 
\\
        \> $|$  \> $(ys = @filter@~f~xs)$    
                \> $=$ \> $C(\Gamma(xs))$ 
\\
        \> $|$  \> $(ys = @generate@~s~f)$  
                \> $=$ \> $C(\Gamma(ys))$ 
\\
        \> $|$  \> $(ys = @gather@~is~xs)$    
                \> $=$ \> $C(\Gamma(is))$ 
\\
        \> $|$  \> $(ys = @cross@~as~bs)$     
                \> $=$ \> $C(\Gamma(as)) \times C(\Gamma(bs))$ 
\\
        \> $|$  \> $(ys = @external@~\overline{xs})$  
                \> $=$ \> $\bot$ 
\\
\end{tabbing}

% Once the constraints are solved, known to be valid, and sorted into equivalence classes, each combinator is assigned a size. Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$. This is because, in order to produce the filtered output, all elements of the input $n$ must be considered.

% After the generated equality constraints are solved, and sizes are grouped into equivalence classes, combinators with iteration sizes in the same equivalence class may be fused together if there are no fusion-preventing dependencies between them.
